{\LARGE Basis Choice for Scalar-on-Function Regression \\ with Applications to Near-Infrared Spectroscopy}

### 2022-02-09 Paper/main.tex
Changes before meeting

	This setup is opposing to the often used penalized functional regression as described by \cite{Goldsmith_2011} in which an explicit smoothness constraint $\lambda$ is used to tune the smoothness of the estimator $\hat{\beta}(t)$ while setting the $K$ sufficiently high. This would avoid the heavy computing of validating the best value for $K$, which we will conduct in the simulation. To provide intuition in this approach, let 
	 \begin{equation}
	 	PSSE_\lambda(\alpha, \beta) = \sum_{i = 1}^{N} \left[ Y_i -\alpha -\int_0^1 \beta(t)X_i(t)dt \right]^2 + \lambda \int \left[D^m\beta(t)\right]^2 dt
	 \end{equation}

	 denote the penalized residual sum of squares as notated by \cite{ramsay_functional_2005} for the derivative of order $m$. A typical choice is the second derivative as highly variable functions are expected to yield large second derivatives and therefore a larger penalty. The smoothing parameter $\lambda$ is set to minimize the $PSSE_\lambda(\alpha, \beta)$, which can be archived by different criteria as shown in \cite{ThomasLee_2003}.

@@

	\subsection{Generating Similar Curves}
	To avoid small sample problems, we generated 200 similar curves, $NIR_{sim}$, from the spectra of the gasoline data set, $NIR$, motivated by \hyperlink{KL}{Karhunen-Lo\'{e}ve Expansion}. First, the initial curves are expressed in terms of a generated cubic B-spline basis which is created using 50 knots. In the R implementation of the fda package that is used, these 50 knots account for 52 basis functions (50+4-2). These smooth curves are then centered, before applying the \hyperlink{KL}{Karhunen-Lo\'{e}ve Expansion}. It is assumed that the scores follow a normal distribution, so the new realizations for the scores are drawn from a multivariate normal distribution $\mathring{\xi} = \left(\mathring{\xi}_{1},\: \dots \:, \mathring{\xi}_{M}\right)' \sim \mathcal{N}(0_M, \; diag(\hat{\lambda}_1,\: \dots\:, \hat{\lambda}_M))$. Finally, we obtain the generated curves $NIR_{sim}$

@@

		\begin{figure}
			\centering
			\begin{minipage}{.5\textwidth}
				\centering
  				\includegraphics[width=\textwidth]{../Graphics/f1_plot.pdf}
  				\caption{$f_1(t)$, smooth function}
  				\label{fig:test1}
			\end{minipage}%
			\begin{minipage}{.5\textwidth}
	  			\centering
  				\includegraphics[width=\textwidth]{../Graphics/f2_plot.pdf}
  				\caption{$f_2(t)$, bumpy function}
  				\label{fig:test2}
			\end{minipage}
		\end{figure}

@@

		where $Z \sim \mathcal{N}(0,1)$. In total, we created four combinations for the simulations, using the two true coefficient functions and the two sets of responses. These four combinations are then used with a different number of monomial basis functions $ \in \{1,2, \dots, 6\}$, cubic B-spline basis-function $\{5,6,...,18\}$ and Fourier functions $\{1,3,...,25\}$ to predict the generated responses using the basis expansion approach and the FPCR approach. For the evaluation, we used the prediction MSE calculated by 10 fold cross-validation.
		To obtain valid out of sample properties for the FPCR, within each of the ten ten-fold cross-validation splits, we first calculate the first $nharm$ FPC's for $nharm \in \{2,3,4\}$ of the training set $\mathcal{T}$ for each curve $i \in \{1,2, \dots, 200\}$, which was smoothed with the respective basis function specification. The obtained Eigenfunctions $\nu^{m,\mathcal{T}}$ are then used to estimate the scores of the holdout set $\mathcal{H}$, $\hat{\xi}_{i}^{m, \mathcal{H}}$  by the equation
		\begin{equation}
			\begin{split}
				\hat{\xi}_{i}^{m, \mathcal{H},L} &\approx  \int_{0}^{1} {\color{red} \left(X_{i}^{\mathcal{H}}(s) - \hat{\mu}^{\mathcal{T}}(s)\right)} {\color{blue}\hat{\nu}^{m, \mathcal{T}} }\mathrm{d}s) 
			    = \int_{0}^{1} {\color{red}\left(\sum_{j = 1}^{L} a_{i,j}^{\mathcal{H}} \phi_j(s)\right)} {\color{blue}\left(\sum_{k = 1}^{L} b_{k}^{m, \mathcal{T}} \phi_{k}(s)\right)} \mathrm{d}s \\
				 &\approx \sum_{j = 1}^{L} \left[ a_{i,j}^{\mathcal{H}}\sum_{k = 1}^{L}  b_{k}^{m, \mathcal{T}} \int_{0}^{1} \phi_j(s) \psi_{k}(s)\mathrm{d}s\right]
			\end{split}
		\end{equation}

		with truncation parameter $L$.
		\vspace{0.2cm}

		The simulation was done with R (version...). In total, 5000 repetitions were done for each set of simulations.

@@

	\section{Application}\label{Application}
		The application uses the methods and insights from the previous sections to predict the octane ratings of the introduced gasoline data set.
		Although the simulation study granted valuable insights in the different methods in four different settings, it is not enough to determine the method and choice of basis of the application, since there is too much uncertainty involved. To point out some sources of uncertainty: First, opposing to the simulation setup, we do not know the true coefficient function. Visual inspection fuels the hypothesis that the estimated coefficient functions for $NIR$ look closer to $f_2$ than to $f_1$ but the insights from two functions are not sufficient to draw any conclusion. Second, we have no information about the signal-to-noise ratios of the measured octane numbers. Third, to generate similar curves, we made assumptions about the distribution of $\mathring{\xi}$ that are not relatable to this real-life application where we not know this distribution. 
		Therefore, we will run all specification again for the gasoline data set and use the results of the simulation study towards the reading of the results. The application is designed similar to the simulation study: The 60 spectra of the gasoline data set will be used to regress them on the reported octane numbers and evaluate the results as prediction MSE using 12 fold cross-validation with 5 elements per fold. In total, we conducted 1000 repetitions for each setting.

	\subsection{Interpretation of Results}
		\subsubsection{Basis Expansion Regression}
		The cross-validated MSPE selects 5 basis functions for the monomial basis. For B-Splines, the cross-validation selects 10 basis functions and 9 basis functions for the fourier basis. Comparing the estimates in \hyperref[basis_expansion_estimate_appl]{ADD}, apart from the boundaries, B-spline and Fourier basis show similar behavior, opposing to the monomial estimate which might be attributed by the lower amount of used basis functions. Especially at the lower boundary, the monomial basis shows a strong exaggerated bahavior. However, we must exercise caution since from $L$=6 onwards, we were not able to calculate stable results for the monomial basis. The MSPE for B-splines (0.04574) and Fourier (0.04808) are similar while it is clearly knocked of for the monomial basis (0.24181). In contrast to the simulation study, the B-splines basis manages to outperform the fourier basis. Driving factors for this better performance could be that first, we do not assume a periodic true coefficient function with the same start- and end value and second, the Fourier basis tries to force the same start- and endvalue on $NIR$ (confirmed in unreported visual inspection). It is worthy to note that the reported MSPE for the B-spline and Fourier basis are close to the errors reported in the simulation study for the setup $f_2,Y_1$, which might be caused by the mentioned similarities between $NIR$ and $f_2$, but also by a similar high signal-to-noise ratio of the octane numbers.

		\subsubsection{Functional Principal Component Regression}
		The best performance for all numbers of FPC's was achieved with the Fourier Basis, as for the FPCR in the simulation study. In contrast to the simulation study, no evidence of overfitting was found for the best basis choices in any of the three bases. The MSPE appears to be strictly decreasing in the number of FPC´s for its best basis choice. As for the simulation study, the interpretation of the results with respect to the chosen basis is difficult: Referring to the plotted estimates $\hat{\beta}(t)$ in \hyperref[FPCR2_estimate_appl]{ADD}, \hyperref[FPCR3_estimate_appl]{ADD} and \hyperref[FPCR4_estimate_appl]{ADD}, it appears that the higher $n_{FPC}$ is, the more similar behaviour of the estimate $\hat{\beta}(t)$ can be observed. Opposing to the simulation study, the number of basis functions is steadily increasing for the monomial basis. The behavior for the B-spline basis is similar to the one reported in the simulation study (basis functions increasing in $n_{FPC}$).

@@

	\subsection{Application - Coefficient Function Estimates}
	\begin{figure}[H]
		\centering
		\begin{minipage}{.5\textwidth}
			\centering
			\includegraphics[width=\textwidth]{../Graphics/Appl\_Curve\_Estimates/appl\_basis\_expansion.pdf}
			\caption{Basis Expansion Regression}
			\label{basis_expansion_estimate_appl}
		\end{minipage}%
		\begin{minipage}{.5\textwidth}
			\centering
			\includegraphics[width=\textwidth]{../Graphics/Appl\_Curve\_Estimates/fpcr\_nharm2.pdf}
			\caption{2 Functional Principal Components}
			\label{FPCR2_estimate_appl}			
		\end{minipage}
	\end{figure}

>>>>

	This setup differs from the often used penalized functional regression as described by \cite{Goldsmith_2011} in which an explicit smoothness constraint is used to tune the smoothness of the estimator $\hat{\beta}(t)$ while setting the $K$ sufficiently high. This would avoid the heavy computations of validating the best value for $K$, which we will conduct in the simulation. To provide intuition in this approach, let 
	 \begin{equation}
	 	PSSE_\lambda(\alpha, \beta) = \sum_{i = 1}^{N} \left[ Y_i -\alpha -\int_0^1 \beta(t)X_i(t)dt \right]^2 + \lambda \int \left[D^m\beta(t)\right]^2 dt
	 \end{equation}

	 denote the penalized residual sum of squares as notated by \cite{ramsay_functional_2005} for the derivative of order $m$. A typical choice is the second derivative as highly variable functions are expected to exhibit large second derivatives and therefore a larger penalty. The smoothing parameter $\lambda$ is set to minimize the $PSSE_\lambda(\alpha, \beta)$, which can be archived by different criteria as shown in \cite{ThomasLee_2003}.

@@

	\subsection{Generating Similar Curves}
	To avoid small sample problems, we generated 200 similar curves, $NIR_{sim}$, from the spectra of the gasoline data set, $NIR$, motivated by \hyperlink{KL}{Karhunen-Lo\'{e}ve Expansion}. First, the initial curves are expressed in terms of a  cubic B-spline basis which is created using 50 knots. In the R implementation of the fda package, these 50 knots account for 52 basis functions (50+4-2). These smooth curves are then centered, before applying the \hyperlink{KL}{Karhunen-Lo\'{e}ve Expansion}. We assume that the scores follow a normal distribution, so the new realizations for the scores are drawn from a multivariate normal distribution $\mathring{\xi} = \left(\mathring{\xi}_{1},\: \dots \:, \mathring{\xi}_{M}\right)' \sim \mathcal{N}(0_M, \; diag(\hat{\lambda}_1,\: \dots\:, \hat{\lambda}_M))$. Finally, we obtain the generated curves $NIR_{sim}$ as

@@

		\begin{figure}[H]
			\centering
			\begin{minipage}{.5\textwidth}
				\centering
  				\includegraphics[width=\textwidth]{../Graphics/f1_plot.pdf}
  				\caption{$f_1(t)$, smooth function}
  				\label{fig:test1}
			\end{minipage}%
			\begin{minipage}{.5\textwidth}
	  			\centering
  				\includegraphics[width=\textwidth]{../Graphics/f2_plot.pdf}
  				\caption{$f_2(t)$, bumpy function}
  				\label{fig:test2}
			\end{minipage}
		\end{figure}

@@

		where $Z \sim \mathcal{N}(0,1)$. In total, we created four combinations, using the two true coefficient functions and the two error terms. These four combinations are then used with a different number of monomial basis functions $ \in \{1,2, \dots, 6\}$, cubic B-spline basis-function $\{5,6,...,18\}$ and Fourier functions $\{1,3,...,25\}$ to predict the generated responses using the basis expansion approach and the FPCR approach. For the evaluation, we used the prediction MSE calculated by 10 fold cross-validation.
		To obtain valid out of sample properties for the FPCR, within each of the ten ten-fold cross-validation splits, we first calculate the first $nharm$ FPC's for $nharm \in \{2,3,4\}$ of the training set $\mathcal{T}$, which was smoothed with the respective basis function specification. The obtained Eigenfunctions $\nu^{m,\mathcal{T}}$ are then used to estimate the scores of the holdout set $\mathcal{H}$, $\hat{\xi}_{i}^{m, \mathcal{H}}$  by the equation
		\begin{equation}
			\begin{split}
				\hat{\xi}_{i}^{m, \mathcal{H},L} &\approx  \int_{0}^{1} {\color{red} \left(X_{i}^{\mathcal{H}}(s) - \hat{\mu}^{\mathcal{T}}(s)\right)} {\color{blue}\hat{\nu}^{m, \mathcal{T}} }\mathrm{d}s) 
			    = \int_{0}^{1} {\color{red}\left(\sum_{j = 1}^{L} a_{i,j}^{\mathcal{H}} \phi_j(s)\right)} {\color{blue}\left(\sum_{k = 1}^{L} b_{k}^{m, \mathcal{T}} \phi_{k}(s)\right)} \mathrm{d}s \\
				 &\approx \sum_{j = 1}^{L} \left[ a_{i,j}^{\mathcal{H}}\sum_{k = 1}^{L}  b_{k}^{m, \mathcal{T}} \int_{0}^{1} \phi_j(s) \psi_{k}(s)\mathrm{d}s\right]
			\end{split}
		\end{equation}

		with truncation parameter $L$. In total, 5000 repetitions were conducted for each set of simulations.

@@

	\section{Application}\label{Application}
		Our application uses the methods and insights from the previous sections to predict the octane ratings of the introduced gasoline data set.
		Although the simulation study granted valuable insights into the different methods in our four different settings, it is not enough to determine the optimal method and choice of basis for the application, since there is too much uncertainty involved. To point out some sources of uncertainty: First, differing from the simulation setup, we do not know the true coefficient function. Visual inspection of the the estimated coefficient functions shwon in Appendix \ref{Estimates_Appl} fuels the hypothesis the real coefficient function might be closer to $f_2$ than to $f_1$ but the insights from two functions are not sufficient to draw any conclusion. Second, we have no information about the signal-to-noise ratios of the measured octane numbers. Third, to generate similar curves, we made assumptions about the distribution of $\mathring{\xi}$ that are not applicable to this real-life application where we do not know their distribution.

		Therefore, we will run all specification again for the gasoline data set and use the results of the simulation study to improve our understanding of the results. The application is designed similar to the simulation study: The 60 spectra of the gasoline data set will be used as regressors to estimate a coefficient function, predict the reported octane numbers and evaluate the results via prediction MSE using 12 fold cross-validation with 5 elements per fold. In total, we conducted 1000 repetitions for each setting.

	\subsection{Interpretation of Results}
		\subsubsection{Basis Expansion Regression}
		The cross-validated MSPE selects 5 basis functions for the monomial basis. For B-Splines, the cross-validation selects 10 basis functions and 9 basis functions for the Fourier basis. Comparing the estimates in Appendix \ref{Estimates_Appl}, apart from the boundaries, the B-spline and Fourier basis show similar behavior, opposing to the monomial estimate which might be attributable to the lower number of basis functions. Especially at the lower boundary, the monomial basis shows an exaggerated behavior. However, we must exercise caution since from $L$=6 onwards, we were not able to calculate stable results for the monomial basis. The MSPE for B-splines (0.04574) and Fourier (0.04808) are similar while the monomial basis (0.24181) shows distinctly worse properties. In contrast to the simulation study, the B-splines basis manages to outperform the Fourier basis. Driving factors for this better performance could be that, first, we do not assume a periodic true coefficient function with the same start- and end value and second, the Fourier basis tries to force the same start- and endvalue on $NIR$. It is worthy to note that the reported MSPE for the B-spline and Fourier basis are close to the errors reported in the simulation study for the setup $f_2,Y_1$, which might be caused by the hypothesized similarities between the true coefficient function and $f_2$, but also by a similar high signal-to-noise ratio of the octane numbers.

		\subsubsection{Functional Principal Component Regression}
		The best performance for all numbers of FPC's was achieved with the Fourier Basis, as for the FPCR in the simulation study. In contrast to the simulation study, no evidence of overfitting was found for the best basis choices in any of the three bases. The MSPE appears to be strictly decreasing in the number of FPC´s for its best basis choice. As for the simulation study, the interpretation of the results with respect to the chosen basis is difficult: Referring to the plotted estimates $\hat{\beta}(t)$ in \hyperref[FPCR2_estimate_appl]{ADD}, \hyperref[FPCR3_estimate_appl]{ADD} and \hyperref[FPCR4_estimate_appl]{ADD}, it appears that the higher $n_{FPC}$ is, the more similar behavior of the estimate $\hat{\beta}(t)$ can be observed. Differing from the simulation study, the number of basis functions is steadily increasing for the monomial basis. The behavior for the B-spline basis is similar to the one reported in the simulation study (basis functions increasing in $n_{FPC}$).

@@

	\subsection{Application - Coefficient Function Estimates}\label{Estimates_Appl}
	\begin{figure}[H]
		\centering
		\begin{minipage}{.5\textwidth}
			\centering
			\includegraphics[width=\textwidth]{../Graphics/Appl\_Curve\_Estimates/appl\_basis\_expansion.pdf}
			\caption{Basis Expansion Regression}
			\label{basis_expansion_estimate_appl}
		\end{minipage}%
		\begin{minipage}{.5\textwidth}
			\centering
			\includegraphics[width=\textwidth]{../Graphics/Appl\_Curve\_Estimates/fpcr\_nharm2.pdf}
			\caption{2 Functional Principal Components}
			\label{FPCR2_estimate_appl}			
		\end{minipage}
	\end{figure}
